---
title: 'Module 7: Lesson 2 Exercises'
author: "Callum Arnold"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up

## Loading packages and data

```{r}
library(doParallel)
library(doRNG)
library(tidyverse)
library(pomp)

registerDoParallel()
registerDoRNG(2488820)
```

```{r}
meas <- read_csv(paste0(
  "https://kingaa.github.io/sbied/stochsim/",
  "Measles_Consett_1948.csv"
  )) %>%
  select(week, reports = cases)
```

```{r}
plot(meas, type = "o")
```

## Setting up basic SIR model

As we've already gone through the process and intuition of setting up the model,
we'll just go straight to the C code.

```{r}
sir_step <- Csnippet("
  double dN_SI = rbinom(S, 1 - exp(-Beta * (I/N) * dt));
  double dN_IR = rbinom(I, 1 - exp(-mu_IR * dt));
  S -= dN_SI;
  I += dN_SI - dN_IR;
  R += dN_IR;
  H += dN_IR;
  ")
```

```{r}
sir_rinit <- Csnippet("
  S = nearbyint(eta * N);
  I = 1;
  R = nearbyint((1 - eta) * N);
  H = 0;
")
```

```{r}
sir_dmeas <- Csnippet("
  lik = dnbinom_mu(reports, k, rho * H, give_log);
  ")

sir_rmeas <- Csnippet("
  reports = rnbinom_mu(k, rho * H);
  ")
```

```{r}
measSIR <- meas %>% 
  pomp(
    times = "week", t0 = 0,
    rprocess = euler(sir_step, delta.t = 1/7),
    rinit = sir_rinit,
    rmeasure = sir_rmeas,
    dmeasure = sir_dmeas,
    accumvars = "H",
    statenames = c("S", "I", "R", "H"),
    paramnames = c("Beta", "mu_IR", "N", "eta", "rho", "k")
    )
```

```{r}
sims <- measSIR %>%
  simulate(
    params = c(
      Beta = 7.5, mu_IR = 0.5, rho = 0.5, k = 10,
      eta = 0.03, N = 38000
      ),
    nsim = 20, format = "data.frame", include.data = TRUE
  )
```

```{r}
sims %>%
  ggplot(aes(x = week, y = reports, group = .id, color = .id == "data")) +
  geom_line() +
  guides(color = FALSE)
```

# Exercise 2.3

> Fiddle with the parameters to see if you can't find a model for which the data
> are a more plausible realization.

## Modifying parameters

### One at a time

To start with, we'll just modify one parameter at a time, starting from the
initial parameter estimates

#### Transmission rate ($\beta$)

```{r}
beta_changes <- purrr::map(
  .x = seq(5, 50, by = 5),
  .f = function(.x){
    sims <- measSIR %>%
      simulate(
        params = c(
        Beta = .x, mu_IR = 0.5, rho = 0.5, k = 10,
        eta = 0.03, N = 38000
          ),
        nsim = 20, format = "data.frame", include.data = TRUE
      )
    
    plot <- sims %>% 
      ggplot(aes(x = week, y = reports, group = .id, color = .id == "data")) +
        geom_line() +
        labs(caption = glue::glue("Beta value = {.x}")) +
        guides(color = FALSE)
    
  }
)
```

```{r}
beta_changes
```

#### Recovery rate ($\mu_{IR}$)

```{r}
mu_ir_changes <- purrr::map(
  .x = seq(0.05, 0.7, by = 0.05),
  .f = function(.x){
    sims <- measSIR %>%
      simulate(
        params = c(
        Beta = 7.5, mu_IR = .x, rho = 0.5, k = 10,
        eta = 0.03, N = 38000
          ),
        nsim = 20, format = "data.frame", include.data = TRUE
      )
    
    plot <- sims %>% 
      ggplot(aes(x = week, y = reports, group = .id, color = .id == "data")) +
        geom_line() +
        labs(caption = glue::glue("Mu (IR) value = {.x}")) +
        guides(color = FALSE)
    
  }
)
```

```{r}
mu_ir_changes
```

#### Over-dispersion parameter ($k$)

```{r}
k_changes <- purrr::map(
  .x = seq(1, 100, by = 10),
  .f = function(.x){
    sims <- measSIR %>%
      simulate(
        params = c(
        Beta = 7.5, mu_IR = 0.5, rho = 0.5, k = .x,
        eta = 0.03, N = 38000
          ),
        nsim = 20, format = "data.frame", include.data = TRUE
      )
    
    plot <- sims %>% 
      ggplot(aes(x = week, y = reports, group = .id, color = .id == "data")) +
        geom_line() +
        labs(caption = glue::glue("k value = {.x}")) +
        guides(color = FALSE)
    
  }
)
```

```{r}
k_changes
```

#### Proportion between observed and actual cases ($\rho$)

```{r}
rho_changes <- purrr::map(
  .x = seq(0.1, 0.9, by = 0.1),
  .f = function(.x){
    sims <- measSIR %>%
      simulate(
        params = c(
        Beta = 7.5, mu_IR = 0.5, rho = .x, k = 10,
        eta = 0.03, N = 38000
          ),
        nsim = 20, format = "data.frame", include.data = TRUE
      )
    
    plot <- sims %>% 
      ggplot(aes(x = week, y = reports, group = .id, color = .id == "data")) +
        geom_line() +
        labs(caption = glue::glue("Rho value = {.x}")) +
        guides(color = FALSE)
    
  }
)
```

```{r}
rho_changes
```

#### Proportion susceptible at start ($\eta$)

```{r}
eta_changes <- purrr::map(
  .x = seq(0.01, 0.2, by = 0.01),
  .f = function(.x){
    sims <- measSIR %>%
      simulate(
        params = c(
        Beta = 7.5, mu_IR = 0.5, rho = 0.5, k = 10,
        eta = .x, N = 38000
          ),
        nsim = 20, format = "data.frame", include.data = TRUE
      )
    
    plot <- sims %>% 
      ggplot(aes(x = week, y = reports, group = .id, color = .id == "data")) +
        geom_line() +
        labs(caption = glue::glue("Eta value = {.x}")) +
        guides(color = FALSE)
    
  }
)
```

```{r}
eta_changes
```

### Multiple parameters

```{r}
betas <- seq(25, 40, 5)
rhos <- 0.5
mus <- seq(0.1, 0.3, 0.05)
etas <- seq(0.05, 0.1, 0.01)
ks <- 10


parameters <- expand_grid(
  betas,
  rhos,
  mus,
  etas,
  ks
)
```

```{r}
parameters
```

```{r}
multiple_parms <- parameters %>% 
  rowwise() %>% 
  mutate(
    models = list(measSIR %>%
      simulate(
        params = c(
          Beta = betas, mu_IR = mus, rho = rhos, k = ks,
          eta = etas, N = 38000
          ),
        nsim = 20, format = "data.frame", include.data = TRUE
      )
  ),
  parms = glue::glue("Beta = {betas}, Mu = {mus}, rho = {rhos}, \nk = {ks}, eta = {etas}")) %>% 
  ungroup() %>% 
  mutate(
    plots = purrr::map2(
      .x = models,
      .y = parms,
      .f = function(.x, .y){
        ggplot(
          data = .x,
          aes(x = week, y = reports, group = .id, color = .id == "data")
          ) +
          geom_line() +
          labs(caption = .y) +
          guides(color = FALSE)
      }
    )
  )
```

```{r}
multiple_parms$plots
```

# Exercise 2.4

> Below is a diagram of the so-called SEIR model. This differs from the SIR
> model in that infected individuals must pass a period of latency before
> becoming infectious.
>
> Modify the codes above to construct a pomp object containing the Consett
> measles data and an SEIR model. Perform simulations as above and adjust
> parameters to get a sense of whether improvement is possible by including a
> latent period.

## Example Code

```{r}
seir_step <- Csnippet("
  double dN_SE = rbinom(S,1-exp(-Beta*I/N*dt));
  double dN_EI = rbinom(E,1-exp(-mu_EI*dt));
  double dN_IR = rbinom(I,1-exp(-mu_IR*dt));
  S -= dN_SE;
  E += dN_SE - dN_EI;
  I += dN_EI - dN_IR;
  R += dN_IR;
  H += dN_IR;
")


seir_init <- Csnippet("
  S = nearbyint(eta*N);
  E = 0;
  I = 1;
  R = nearbyint((1-eta)*N);
  H = 0;
")

measSIR %>%
  pomp(
    rprocess=euler(seir_step,delta.t=1/7),
    rinit=seir_init,
    paramnames=c("N","Beta","mu_EI","mu_IR","rho","eta"),
    statenames=c("S","E","I","R","H")
  ) -> measSEIR
```

```{r}
measSEIR %>%
  simulate(params=c(Beta=30,mu_EI=0.8,mu_IR=1.3,rho=0.5,k=10,eta=0.06,N=38000),
    nsim=20,format="data.frame",include.data=TRUE) %>%
  ggplot(aes(x=week,y=reports,group=.id,color=.id=="data"))+
  geom_line()+
  guides(color="none")
```

```{r}
measSEIR %>% 
  simulate(params=c(Beta=40,mu_EI=0.8,mu_IR=1.3,rho=0.5,k=10,eta=0.06,N=38000),
  nsim=20,format="data.frame",include.data=TRUE) %>%
  ggplot(aes(x=week,y=reports,group=.id,color=.id=="data"))+
  geom_line()+
  guides(color="none")
```

## Group code

```{r}
#Add exposed function
#Added Exposed compartment 
# Added mu_EI for rate from Exposed to infectious
sir_step <- function (S, E, I, R, N, Beta, zeta_EI, mu_IR, delta.t, ...) {
  
  dN_SE <- rbinom(n=1,size=S,prob=1-exp(-Beta*I/N*delta.t))
  
  #Add equation to describe movement form Exposed to infectious
  dN_EI <- rbinom(n=1, size=E, prob = 1 - exp(-zeta_EI*detla.t))
  dN_IR <- rbinom(n=1,size=I,prob=1-exp(-mu_IR*delta.t))
  
  S <- S - dN_SE
  E <- E + dN_SE - dN_EI
  I <- I + dN_EI - dN_IR
  R <- R + dN_IR
  c(S = S, E = E, I = I, R = R)
}

sir_rinit <- function (N, eta, ...) {
  c(S = round(N*eta), I = 1, R = round(N*(1-eta)))
}

#Work with Pomp. 
meas %>%
  pomp(times="week",t0=0,
       rprocess=euler(sir_step,delta.t=1/7),
       rinit=sir_rinit
  ) -> measSIR

sir_step <- function (S, E, I, R, H, N, Beta, mu_IR, zeta_EI, delta.t, ...)
{
  dN_SI <- rbinom(n=1,size=S,prob=1-exp(-Beta*I/N*delta.t))
  #Add equation to describe movement form Exposed to infectious
  dN_EI <- rbinom(n=1, size=E, prob = 1 - exp(-zeta_EI*detla.t))
  dN_IR <- rbinom(n=1,size=I,prob=1-exp(-mu_IR*delta.t))
  S <- S - dN_SE
  E <- E + dN_SE - dN_EI
  I <- I + dN_EI - dN_IR
  R <- R + dN_IR
  H <- H + dN_IR;
  c(S = S, E = E, I = I, R = R, H = H)
}

sir_rinit <- function (N, eta, ...) {
  c(S = round(N*eta), I = 1, R = round(N*(1-eta)), H = 0)
}

measSIR %>% 
  pomp(
    rprocess=euler(sir_step,delta.t=1/7),
    rinit=sir_rinit, accumvars="H"
  ) -> measSIR

sir_dmeas <- function (reports, H, rho, k, log, ...) {
  dnbinom(x=reports, size=k, mu=rho*H, log=log)
}

sir_rmeas <- function (H, rho, k, ...) {
  c(reports=rnbinom(n=1, size=k, mu=rho*H))
}

measSIR %>%
  pomp(
    rmeasure=sir_rmeas,
    dmeasure=sir_dmeas
  ) -> measSIR



sir_step <- Csnippet("
  double dN_SE = rbinom(S,1-exp(-Beta*I/N*dt));
  double dN_EI = rbinom(S,1-exp(-*I/N*dt));
  double dN_IR = rbinom(I,1-exp(-mu_IR*dt));
  S -= dN_SI;
  I += dN_SI - dN_IR;
  R += dN_IR;
  H += dN_IR;
  ")

sir_rinit <- Csnippet("
  S = nearbyint(eta*N);
  I = 1;
  R = nearbyint((1-eta)*N);
  H = 0;
  ")

sir_dmeas <- Csnippet("
  lik = dnbinom_mu(reports,k,rho*H,give_log);
  ")

sir_rmeas <- Csnippet("
  reports = rnbinom_mu(k,rho*H);
  ")

measSIR %>%
  pomp(rprocess=euler(sir_step,delta.t=1/7),
       rinit=sir_rinit,
       rmeasure=sir_rmeas,
       dmeasure=sir_dmeas,
       accumvars="H",
       statenames=c("S","I","R","H"),
       paramnames=c("Beta","mu_IR","N","eta","rho","k")
  ) -> measSIR
```
